---
title: "KODAQs Assignment"
author: "Elnari Potgieter"
date: "02/15/2026"
format: 
  html:
    fig-width: 8
    fig-height: 4
    code-fold: true
---

# English Monarchs and Marriages - A Data Story

## Introduction

This report uses a data set from GitHub's Tidy Tuesdays called "English Monarchs and Marriages" to determine the average age of marriage of English monarchs and their consorts at the time of marriage, as well as the oldest and youngest ages of both monarchs and consorts at the time of marriage.

The link for the data set can be found here: https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-08-20/english_monarchs_marriages_df.csv.

## Loading Libraries and Data

Firstly, I load the libraries I need for my analysis.

```{r}
#| include: false
library (dplyr)
library (psych)
library(ggplot2)
library (knitr)
```

As I do not know the data well, I get an overview of the variables and structure.

```{r}
# Load data from Github
english_monarchs_marriages_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-08-20/english_monarchs_marriages_df.csv')

# Check variables and data structure
summary(english_monarchs_marriages_df)
str(english_monarchs_marriages_df)

```

From the above, I can see that there are five variables: king_name, king_age, consort_name, consort_age and marriage_year. I also can see that there are 83 rows, assuming that n = 83. Some monarchs and consorts seem to have married more than once, but this is not a problem as I am interested in ages when married (whether once or more).

## Cleaning and Preparing Data

I notice that there are some missing values indicated by "?", "-" or "NA". I also notice that some values are not certain, indicated with "(?)". I proceed to create a clean data set by removing the rows with incomplete data, but keep the uncertain dates as best estimates.

```{r}

# Ensure that variables are treated as characters to filter out (?)
english_monarchs_marriages_df$king_age <- as.character(english_monarchs_marriages_df$king_age)

english_monarchs_marriages_df$consort_age <- as.character(english_monarchs_marriages_df$consort_age)

english_monarchs_marriages_df$year_of_marriage <- as.character(english_monarchs_marriages_df$year_of_marriage)


# Remove (?) from uncertain dates and change variable names to shorter / more sensible (some queens among the king list)


emm_clean <- english_monarchs_marriages_df %>%
  mutate(ruler_age = gsub("\\(?)", "", king_age)) %>%
  mutate(consort_age = gsub("\\(?)", "", consort_age)) %>%
  mutate(marriage_year = gsub("\\(?)", "", year_of_marriage))

# Render rows with ? or - missing in age or year related columns

columns_to_clean <- c("ruler_age", "consort_age", "marriage_year")

emm_clean[columns_to_clean] <- lapply(emm_clean[columns_to_clean], as.character)


emm_clean[columns_to_clean] <- lapply(emm_clean[columns_to_clean], function(x) {
  x[x == "–"] <- NA
  x[x == "?"] <- NA
  return(x)
})

# Remove rows with NA 

emm_clean <- na.omit(emm_clean)

# (?) not entirely removed, manually replace with number

emm_clean$ruler_age <- replace(emm_clean$ruler_age, emm_clean$ruler_age == "50(?", "50")

emm_clean$marriage_year <- replace(emm_clean$marriage_year, emm_clean$marriage_year == "1044(?", "1044")

# Rename king_name to align with ruler_name

emm_clean <- emm_clean %>%
  rename(ruler_name = king_name) 


# Create final data set with columns in order

emm_final <- emm_clean %>%
  select (ruler_name, ruler_age, consort_name, consort_age, marriage_year)

# Check dataset
summary (emm_final)

```

Following the removal of missing values, there is some data loss with n now at 58.

# Analysis and Findings

### Ages of Monarchs and Their Consorts

This sections determines the oldest and youngest monarch and consort age, as well as the average age of monarchs and consorts at the time of marriage. In addition, I determine who the youngest and oldest monarchs and consorts at the time of their marriage were.

```{r}
# Make sure that relevant variables are treated as numeric

emm_final$ruler_age <- as.numeric(emm_final$ruler_age)
emm_final$consort_age <- as.numeric(emm_final$consort_age)

# Determine the min, max and average using describe

age_desc_table <- describe(emm_final[, c("ruler_age", "consort_age")])

# Convert the describe() output to a data frame
age_desc_table_df <- as.data.frame(age_desc_table)

# Save the summary table as a CSV file
write.csv(age_desc_table_df, "tables/age__desc_table.csv", row.names = FALSE)

# Display the table using kable
kable(age_desc_table, align = "lcccccc", caption = "Table 1: Descriptive Statistics for Monarch and Consort Ages")

```

From the above, we can determine that the youngest monarch was 5 at the time of marriage (minimum) and the oldest 60 at the time of marriage (maximum), while the youngest consort was 3 at the time of marriage and the oldest 41 at the time of marriage. The average age of marriage for monarchs is 29 (rounded mean) and for consorts 21 (rounded mean), showing that on average monarchs are older than their consorts. The standard deviation for monarch ages (12) show a greater spread in ages for monarchs than the standard deviation (7) for consort ages. With a slightly larger mean than median for monarch ages (29 greater than 26), the results are positively skewed for this variable, as well as a slight positive skew for consort ages (21 greater than 20). The spread and distribution are visualised in the graph below.

```{r}
# Graph showing distribution

# Reshape the data for plotting
plot_data <- data.frame(
  Age = c(emm_final$ruler_age, emm_final$consort_age),
  Type = c(rep("Ruler", length(emm_final$ruler_age)), rep("Consort", length(emm_final$consort_age)))
)
# Create the density plot
age_distr_plot <- ggplot(plot_data, aes(x = Age, fill = Type)) +
  geom_density(alpha = 0.5) +  # Add transparency to distinguish overlapping areas
  labs(title = "Fig 1: Age Distribution of Monarchs and Consorts",
       x = "Age",
       y = "Density") +
  theme_minimal() +
  scale_fill_manual(values = c("Ruler" = "blue", "Consort" = "red"))

# Save the plot as a PDF file
ggsave("figures/age_distr_plot.pdf", plot = age_distr_plot, width = 8, height = 6)


# Display

age_distr_plot
```

The above graph shows that the ages of consorts are more tightly packed around the mean than for monarchs, as well as shows the younger mean for consorts than for monarchs.

Now we determine who the youngest and oldest monarchs and consorts were.

```{r}
# Determine who the youngest and oldest monarchs and consorts were

# Find all rulers with the youngest age
youngest_rulers <- emm_final[which(emm_final$ruler_age == min(emm_final$ruler_age, na.rm = TRUE)), ]
oldest_rulers <- emm_final[which(emm_final$ruler_age == max(emm_final$ruler_age, na.rm = TRUE)), ]
# Find all consorts with the youngest age
youngest_consorts <- emm_final[which(emm_final$consort_age == min(emm_final$consort_age, na.rm = TRUE)), ]
oldest_consorts <- emm_final[which(emm_final$consort_age == max(emm_final$consort_age, na.rm = TRUE)), ]
# Combine the results into a table
result_table <- rbind(
  cbind(Type = "Youngest Ruler", Name = youngest_rulers$ruler_name, Age = youngest_rulers$ruler_age, Marriage_Year = youngest_rulers$marriage_year, Spouse = youngest_rulers$consort_name, Spouse_Age = youngest_rulers$consort_age),
  cbind(Type = "Oldest Ruler", Name = oldest_rulers$ruler_name, Age = oldest_rulers$ruler_age, Marriage_Year = oldest_rulers$marriage_year, Spouse = oldest_rulers$consort_name, Spouse_Age = oldest_rulers$consort_age),
  cbind(Type = "Youngest Consort", Name = youngest_consorts$consort_name, Age = youngest_consorts$consort_age, Marriage_Year = youngest_consorts$marriage_year, Spouse = youngest_consorts$ruler_name, Spouse_Age = youngest_consorts$ruler_age),
  cbind(Type = "Oldest Consort", Name = oldest_consorts$consort_name, Age = oldest_consorts$consort_age, Marriage_Year = oldest_consorts$marriage_year, Spouse = oldest_consorts$ruler_name, Spouse_Age = oldest_consorts$ruler_age)
)


```

Create table with results.

```{r}
# Convert to a data frame for better readability
result_table <- as.data.frame(result_table)

# Save Table as CSV File
write.csv(result_table, "tables/age_table.csv", row.names = FALSE)

# Display the table using kable
kable(result_table, align = "lcccccc", caption = "Table 2: Youngest and Oldest Monarchs, Consorts, Their Spouses and Year of Marriage")

```

From the above, we can see that the youngest couple at the time of their marriage were Henry the Young King, who was 5 years of age when he married Margaret of France - who was 3 years of age at the time of their marriage in the year 1160. The oldest consort at the time of their marriage was Wallis Simpson who married Edward VIII in the year 1937 at the age of 41. The monarch who was the oldest at the time of their marriage was Edward I - 60 years of age at his time of marriage to Margaret of France (presumably a different Margaret of France to the former), who was 20 at the time of their marriage in the year 1299 - indicating a 40 year age gap at the time of their marriage.

### Age Gaps Between Monarchs and Consorts

Following the above, further questions that arise from the large age gap between Edward I and Margaret of France pertain to the spread in age gaps, average age gap, smallest age gap, and largest age gap between monarchs and consorts.

```{r}
# Create a new variable for the absolute age gap
emm_final$age_gap <- abs(emm_final$ruler_age - emm_final$consort_age)

# Ensure the variables is numeric
emm_final$age_gap <- as.numeric(emm_final$age_gap)


# Use the describe() function to summarize the age_gap variable
age_gap_desc_table <- describe(emm_final$age_gap)

# Convert the describe() output to a data frame
age_gap_desc_table_df <- as.data.frame(age_gap_desc_table)

# Save the summary table as a CSV file
write.csv(age_gap_desc_table_df, "tables/age__gap_desc_table.csv", row.names = FALSE)

# Display the table using kable
kable(age_gap_desc_table, align = "lcccccc", caption = "Table 3: Descriptive Statistics for Age Gaps Between Couples")


```

The average age gap between couples is 11 years (mean), with the smallest gap 0 years and the largest gap 40 years. The mean is larger than the median, showing a positive skew. With the minimum at 0 and maximum at 40, the standard deviation of 10 shows some spread in age gaps. This is more explored in the following graph.

```{r}
# Create density plot to show spread in age gaps

age_gap_plot <- ggplot(emm_final, aes(x = age_gap)) +
  geom_density(fill = "green", alpha = 0.5) +
  labs(
    title = "Fig 2: Density Plot of Age Gaps Between Monarchs and Consorts",
    x = "Age Gap",
    y = "Density"
  ) +
  theme_minimal()

# Save the plot as a PDF file

ggsave("figures/age_gap_plot.pdf", plot = age_gap_plot, width = 8, height = 6)

# Display

age_gap_plot

```

The above graph shows the positive skew in age gaps, as well as clustering around the mean tailing off towards larger age gaps. This indicates that large age gaps was not the norm between monarchs and consorts.

I now look at which couples had the largest and smallest age gaps.

```{r}
# Find all rows with the largest age gap
largest_gap_rows <- emm_final[which(emm_final$age_gap == max(emm_final$age_gap, na.rm = TRUE)), ]
# Find all rows with the smallest age gap
smallest_gap_rows <- emm_final[which(emm_final$age_gap == min(emm_final$age_gap, na.rm = TRUE)), ]

# Create a table with the results
gap_table <- rbind(
  cbind(
    Gap_Type = "Largest Gap",
    Ruler_Name = largest_gap_rows$ruler_name,
    Ruler_Age = largest_gap_rows$ruler_age,
    Consort_Name = largest_gap_rows$consort_name,
    Consort_Age = largest_gap_rows$consort_age,
    Marriage_Year = largest_gap_rows$marriage_year,
    Age_Gap = largest_gap_rows$age_gap
  ),
  cbind(
    Gap_Type = "Smallest Gap",
    Ruler_Name = smallest_gap_rows$ruler_name,
    Ruler_Age = smallest_gap_rows$ruler_age,
    Consort_Name = smallest_gap_rows$consort_name,
    Consort_Age = smallest_gap_rows$consort_age,
    Marriage_Year = smallest_gap_rows$marriage_year,
    Age_Gap = smallest_gap_rows$age_gap
  )
)
# Convert to a data frame for better readability
gap_table <- as.data.frame(gap_table)

# Save Table as CSV File
write.csv(gap_table, "tables/gap_table.csv", row.names = FALSE)

# Display the table using kable
kable(gap_table, align = "lcccccc", caption = "Table 4: Monarch Marriages and Largest and Smallest Age Gaps")

```

The couple with the largest age gap is Edward I (also the oldest monarch) and Margaret of France - married in 1299 - with a 40 year age gap. There are two couples with the smallest age gap namely George II and Caroline of Ansbach - married in 1705 - with no age gap (in years, and Victoria and Albert of Saxe-Coburg and Gotha - married in 1840 - also with no age gap.

## Summary and Conclusion

In summary, although the youngest monarch and consort to marry were 5 and 3 years of age respectively, the average age of monarchs and consorts - 29 and 21 respectively - coupled with general clustering of data around the mean, show that marriages that young were not common over the course of history. However, the ages of consorts seemingly were in general younger than the monarchs' ages.

Given the age gap of the oldest monarch at the time of marriage with his consort - a 40 year age gap - coupled with the difference in means between the ages of consorts and monarchs, an analysis of age gaps were done.The largest age gap was 40 years and the smallest 0 years. With a mean of 11 years, the average age gap between monarchs and consorts confirms a general trend of older monarchs marrying younger consorts.

## Disclosure

Use of ChatUiT for coding assistance.
